Gametic Selection, Sex Ratio Bias, and Transitions Between Sex Determination Systems

### 2016-09-06 LaTeX/Turnover.tex
general solutions for W and Y invasions finished

\subsection*{Resident equilibrium and stability}
With allele $M$ fixed, we follow the dynamics of the frequency of $A$ in gametes/gametophytes from the homogametic sex (e.g., in eggs from an XX female, $p_{Xf}$) and the frequencies of $A$ in the two types of gametes/gametophytes from the heterogametic sex (e.g., in sperm from an XY male that are X-bearing, $p_{Xm}$, and Y-bearing, $p_{Ym}$).
Assuming selection is weak relative to recombination, the differences in the frequencies of $A$ in each type of gamete/gametophyte are small, and we can solve for the mean frequency of $A$ across all types ($p_A$) as well as the difference in the frequencies of $A$ between two of the three types, to first order in selection.
Linear stability analysis can then be used to determine the stability of this equilibrium.
With no haploid selection and $k=0$ (neo-$Y$ invading a $ZW$ system) or $k=1$ (neo-$W$ invading an $XY$ system), our results reduce to those of \cite{vanDoorn:2010hu}.

\subsection*{Sex chromosome turnover}
The spread of a rare mutant $m$ at the \textbf{M} locus in such a population is determined by the leading eigenvalue, $\lambda$, of the system described by the equations for the next generation frequency of haploid genotypes with the mutation. %, evaluated at the above equilibrium.
Below we present the results for invasion into an $XY$ system assuming equal dominance coefficients in the two sexes, no competition among gametes/gametophytes from females, and linear arrangement \textbf{X}\textbf{A}\textbf{M}.
Equivalent results for invasions into a $ZW$ system are given in the Sup.\ Mat.

@@

The growth rate of a rare, dominant neo-$Y$ region ($k=0$) is
\begin{equation}
\lambda_{Y,XY} = 1 + V_A \frac{(r - R)}{r R} \frac{s_f^2}{\left( s_f + s_m \right)^2} t_m^2,
\end{equation}
where $V_A = p_A(1-p_A)$ is the variance at the \textbf{A} locus, $s_f$ and $s_m$ are the respective selection coefficients for $A$ in diploid females and males, and $t_m$ is the selection coefficient for $A$ in gametes/gametophytes from males.
The neo-Y can therefore invade whenever it is in tighter linkage with the selected locus than the ancestral sex-determining locus, $r>R$, provided locus \textbf{A} is polymorphic ($V_A>0$) and there is selection among females ($s_f\neq0$) and gametes/gametophytes from males ($t_m\neq0$).
%write in general form without weak selection

@@

The growth rate of a rare, dominant neo-$W$ region ($k=1$) is
\begin{equation}
\lambda_{W,XY} = 1 + V_A \frac{s_f \left( (2 r (1-R) +R) s_f + (1- 2r) R s_m \right)}{2 r R \left( s_f + s_m \right)^2} t_m^2.
\end{equation}
In this case, even when the novel sex-determining locus is in looser linkage with the selected locus than is the ancestral sex-determining locus, $r<R$, a novel sex-determiner can invade.
For example, with $R=1/2$ the neo-W invades if there is some linkage between the ancestral sex-determining and selected loci ($r<1/2$), there is selection among gametes/gametophytes in males ($t_m\neq0$), and there is selection for or against $A$ in both males and females ($s_m s_f >0$) that is stronger in females than in males ($|s_f| > |s_m|$).
%write in general form without weak selection

@@

The growth rate of a rare, dominant neo-ESD region that produces males or females with equal probability ($k=1/2$) is
\begin{equation}
\lambda_{ESD,XY} = 1 + \frac{(\lambda_{Y,XY} - 1) + (\lambda_{W,XY} -1)}{4} \Big|_{R=1/2}
%V_A \frac{r}{1-2r}  \frac{s_{f} \left( s_m - 3 s_f\right) }{8 \left( s_{f} + s_{m} \right)^2} t_{m}^2.
\end{equation}
Thus the $k=1/2$ case gets half of the advantages of a neo-$W$ and half that of a neo-$Y$.  
Recombination between the selected locus and the novel sex-determining locus, $R$, doesn't enter into the $k=1/2$ results because sex is essentially randomized each generation, %as if the novel sex-determiner were on an unlinked site that half the time acts as a neo-$W$ and half the time as a neo-$Y$.
preventing associations from building up between allele $A$ and sex.  
%write in general form without weak selection

>>>>

\subsection*{Resident equilibrium and stability}
With allele $M$ fixed, we follow the dynamics of the frequency of $A$ in gametes/gametophytes from the homogametic sex (e.g., in eggs from an XX female, $p_{Xf}$) and the frequencies of $A$ in the two types of gametes/gametophytes from the heterogametic sex (e.g., in sperm from an XY male that are X-bearing, $p_{Xm}$, and Y-bearing, $p_{Ym}$).
Assuming selection is weak relative to recombination, the differences in the frequencies of $A$ in each type of gamete/gametophyte are small, and we can solve for the mean frequency of $A$ across all types ($p_A$) as well as the difference in the frequencies of $A$ between two of the three types, to first order in selection.
Linear stability analysis can then be used to determine the stability of this equilibrium.
With no haploid selection our results reduce to those of \cite{vanDoorn:2010hu} when $k=0$ (neo-$Y$ invading a $ZW$ system) or when $k=1$ (neo-$W$ invading an $XY$ system).

\subsection*{Sex chromosome turnover}
The spread of a rare mutant $m$ at the \textbf{M} locus in such a population is determined by the leading eigenvalue, $\lambda$, of the system described by the equations for the next generation frequency of haploid genotypes with the mutation. %, evaluated at the above equilibrium.
Below we present the results for invasion into an $XY$ system assuming no competition among gametes/gametophytes from females, and linear arrangement \textbf{X}\textbf{A}\textbf{M}. %equal dominance coefficients in the two sexes,
Equivalent results for invasions into a $ZW$ system are given in the Sup.\ Mat.

@@

%write in general form without weak selection
A rare, dominant neo-$Y$ ($k=0$) is always expected to invade the ancestral $XY$ system when the average growth rate of mutant haplotypes ($XAm$ and $Yam$) is positive, $(g_A + g_a)/2 > 0$ (growth rate is unaffected by the allele at the \textbf{X} locus, but recombination between it and the other loci can still destroy the haplotype). %($g_i = where $\lambda_i$ are the two eigenvalues when $R=r=0$; the \textbf{X} locus does not affect the dynamics).
Defining $g_i^*$ as the growth rate of $Xm$ and $Ym$ haplotypes when on an $A$ ($i=A$) or $a$ background ($i=a$), then $g_i^* < g_i$, and if the $Xm$ and $Ym$ haplotypes decline on both \textbf{A} backgrounds ($g_i^* < 0 \; \forall \; i$) the neo-$Y$ does not invade.
Otherwise, the $Xm$ and $Ym$ haplotypes grow on one \textbf{A} background and decline on the other and neo-$Y$ invasion requires
\begin{equation}
r R \left[ \frac{p_{Xf} w_{m,a}}{g_a^*} + \frac{(1 - p_{Xf}) w_{m,A}}{g_A^*} \right] w_{m,Aa} < \nu_m,
\end{equation}
where $w_{m,i}$ is the relative viability fitness of males depending on their haploid or diploid genotype at the \textbf{A} locus (with $i\in\{A,a,AA,Aa,aa\}$) and
$\nu_m = 
  p_{Xf} p_{Ym} w_{a,m} w_{AA,m} +
  p_{Xf} (1-p_{Ym}) w_{a,m} w_{Aa,m} +
  (1-p_{Xf}) p_{Ym} w_{a,m} w_{Aa,m} +
  (1-p_{Xf}) (1-p_{Ym}) w_{a, m} w_{aa,m}$
is the mean relative fitness of resident males.
%and $g_j^*$ is the growth rate of $Xm$ and $Ym$ haplotypes when on an $A$ ($j=A$)  or $a$ background ($j=a$).
Neo-$Y$ invasion therefore occurs for any recombination rates, $r$ and $R$, when the net flow of double recombinants is from the less fit to the more fit \textbf{A} background (making the term inside the square brackets negative).
%Assuming $g_A^* > 0$ and $g_a^* < 0$ without loss of generality, this is $-(1 - p_{Xf}) w_{m,A} g_a^* < p_{Xf} w_{m,a} g_A^*$.
When the net flow of double recombinants is from the more fit to the less fit haplotype, neo-$Y$ invasion can still occur when the rates of recombination are small enough.

%weak selection
Assuming weak selection, we can solve for the invasion fitness of the neo-$Y$ explicitly, giving
\begin{equation}
\lambda_{Y,XY} \approx 1 + V_A \frac{(r - R)}{r R} \frac{s_f^2}{\left( s_f + s_m \right)^2} t_m^2,
\end{equation}
where $V_A = p_A(1-p_A)$ is the variance at the \textbf{A} locus, $s_f$ and $s_m$ are the respective selection coefficients for $A$ in diploid females and males, and $t_m$ is the selection coefficient for $A$ in gametes/gametophytes from males.
The neo-$Y$ can therefore invade whenever it is in tighter linkage with the selected locus than the ancestral sex-determining locus, $r>R$, provided locus \textbf{A} is polymorphic ($V_A>0$) and there is selection among both females ($s_f\neq0$) and gametes/gametophytes from males ($t_m\neq0$).

@@

%write in general form without weak selection
Similarly, a rare, dominant neo-$W$ ($k=1$) will invade the ancestral $XY$ system whenever the average growth rate of $XAm$ and $Xam$ haplotypes is positive, $(g_A + g_a) / 2 > 1$. 
When the growth rates of mutant haplotypes without recombination ($R=0$) are negative, $g_i^* < 0 \; \forall \; i \in \{A,a\}$, where $g_i < g_i^*$, the neo-$W$ does not invade.
Otherwise neo-$W$ invasion requires
\begin{equation}
R \left[ \frac{\bar{p}_m w_{m,A}}{g_a^*} + \frac{(1 - \bar{p}_m) w_{m,a}}{g_A^*} \right] w_{f,Aa} < \nu_f
\end{equation}
where $\bar{p}_m = (p_{Ym}+p_{Xm}) / 2$ is the mean frequency of $A$ in gametes/gametophytes from males, $w_{f,i}$ is the relative viability fitness of females depending on their diploid genotype at the \textbf{A} locus (with $i\in\{AA,Aa,aa\}$), and 
$\nu_f = 
  p_{Xf} p_{Xm} w_{a,m} w_{AA,f} +
  p_{Xf} (1-p_{Xm}) w_{a,m} w_{Aa,f} +
  (1-p_{Xf}) p_{Xm} w_{a,m} w_{Aa,f} +
  (1-p_{Xf}) (1-p_{Xm}) w_{a, m} w_{aa,f}$
is the mean relative fitness of resident females. 
%The rate of recombination between the \textbf{X} and \textbf{A} loci, $r$, does not explicitly appear because females are homozygous at the \textbf{X} locus with an ancestral $XY$ system.
As in the case of the neo-$Y$, neo-$W$ invasion therefore occurs with any recombination rate, $R$, when the net flow of recombinants is from the less fit to the more fit haplotype.
%Assuming $g_A^*>0$ and $g_a^*<0$ without loss of generality, this is $(1 - \bar{p}_m) w_{m,a} g_a^*< \bar{p}_m w_{m,A} g_A^*$.
And when the net flow of recombinants is from the more fit to the less fit  haplotype, neo-$W$ invasion can still occur when the rate of recombination is small enough.

%weak selection
Assuming weak selection, we can solve for the invasion fitness of the neo-$W$ explicitly, giving
\begin{equation}
\lambda_{W,XY} \approx 1 + V_A \frac{s_f \left( (2 r (1-R) +R) s_f + (1- 2r) R s_m \right)}{2 r R \left( s_f + s_m \right)^2} t_m^2.
\end{equation}
In this case, even when the novel sex-determining locus is in looser linkage with the selected locus than is the ancestral sex-determining locus, $r<R$, a novel sex-determiner can invade.
For example, with $R=1/2$ the neo-$W$ invades if there is any linkage between the ancestral sex-determining and selected loci ($r<1/2$), there is selection among gametes/gametophytes in males ($t_m\neq0$), and there is selection for or against $A$ in both males and females ($s_m s_f >0$) that is stronger in females than in males ($|s_f| > |s_m|$).

@@

The growth rate of a rare, dominant neo-ESD region that produces males or females with equal probability ($k=1/2$) is
\begin{equation}
\lambda_{ESD,XY} \approx 1 + \frac{(\lambda_{Y,XY} - 1) + (\lambda_{W,XY} -1)}{4} \Big|_{R=1/2}
%V_A \frac{r}{1-2r}  \frac{s_{f} \left( s_m - 3 s_f\right) }{8 \left( s_{f} + s_{m} \right)^2} t_{m}^2.
\end{equation}
Thus the $k=1/2$ case gets half of the advantages of a neo-$W$ and half that of a neo-$Y$ (BUT WHY THE 4?).  
Recombination between the selected locus and the novel sex-determining locus, $R$, doesn't enter into the $k=1/2$ results because sex is essentially randomized each generation, %as if the novel sex-determiner were on an unlinked site that half the time acts as a neo-$W$ and half the time as a neo-$Y$.
preventing associations from building up between allele $A$ and sex.  
%write in general form without weak selection
